Type theories as quotient inductive-recursive types

### 2025-02-20 tex/main.tex
Some more appendix proof details

The lifting $\sigma^+$ interacts well with identities and composition.
For comparision, the reader is invited to prove the corresponding QIIT version with transports of the following lemma.
%
\begin{lemma} \label{lem:lifting}
  For all contexts $\Gamma$, $\Delta$, $\Theta$, types $A : \Ty\,\Theta$ and substitutions $\sigma : \Sub\,\Gamma\,\Delta$ and $\tau : \Sub\,\Delta\,\Theta$, we have $\idS^+ =^{\Sub\;(\Theta, A)\;(\Theta, A)} \idS$ and $(\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+$.
\end{lemma}

@@

As we are working with a different type theory from \cref{subsec:SC+U}, we  have to again prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$, but the same argument still applies:
\begin{proposition} \label{prop:correctness-3}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
\end{proposition}

@@

\propConfluenceTwo*
\begin{proof}
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
  \[
    \subTy{(\El\,u)}{\idS}, \qquad \subTy{(\El\,u)}{\sigma;\tau}, \qquad \subTy{(\El\,u)}{\pi_1(\sigma, t)}, \qquad\text{and}\qquad \subTy{(\El\,u)}{\pi_1(\sigma; \tau)}
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El\,(\subTm{\subTm{u}{\tau}}{\sigma}), \qquad \El\,(\subTm{u}{\sigma}), \qquad\text{and}\qquad
    \El\,(\subTm{\subTm{u}{\pi_1\tau}}{\sigma})
  \]
  respectively, no matter which function clause of $\subTy{\blank}{\blank}$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}

@@

\propCoherence*
\begin{proof}
  By \cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
  \begin{align*}
    \subTy{A}{\sigma; (\tau, t)}    & = \subTy{A}{(\sigma;\tau), \sub{t}{\sigma}},
                                     & \subTy{A}{\sigma}               & = \subTy{A}{\emptyctx},
                                     & \subTy{A}{\sigma}               & = \subTy{A}{\pi_1\sigma, \pi_2\sigma}
  \end{align*}
  while other cases hold strictly.
  Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
  However, all cases follow easily from \cref{prop:correctness-1}.
  For example, consider the first case:
  \begin{align*}
    \subTy{(\El\,u)}{\sigma; (\tau, t)} & \judgeq \El(\subTm{u}{\sigma; (\tau, t)}) && \text{by \eqref{eq:def-type-subst-7}} \\
                                      & = \El( \sub{u}{\sigma; (\tau, t)})            && \text{by \cref{prop:correctness-1}} \\
                                      & = \El( \sub{u}{\sigma; \tau, \sub{t}{\sigma}})     && \text{by $\mathsf{concat}$} \\
                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by \cref{prop:correctness-1}} \\
                                      & \judgeq \subTy{\left(\El\,u\right)}{\sigma; \tau, \sub{t}{\sigma}} && \text{by \eqref{eq:def-type-subst-7} }
  \end{align*}
  The remaining two cases are omitted.

  As for term substitution $\subTm{\blank}{\blank}$, each case follows from \cref{prop:correctness-1} in conjunction with the congruence rule for the identity type.
\end{proof}

@@

@@

\propCoherenceUparrow*
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily.
\end{proof}

\coherenceTel*
\begin{proof}
  \Fredrik{TODO: spell out all cases here}
  The proposition is easy to prove for the following equality
  constructors, since the lifting and substitution operations compute
  independently for these cases:
  \[
    \mathsf{idr} : \sigma;\idS = \sigma, \quad
    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma),
  \]
  Take $\mathsf{idr}$ as an example, we can see 
  \begin{align*}
    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows \subTel{\Xi}{\idS}) ; (\idS \upuparrows \Xi) \\
                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
                                   & = \sigma \upuparrows \Xi
  \end{align*}
  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.

@@

The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
  \LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
  For the following equality constructors $\cc : \sigma = \tau$
  \begin{alignat*}{5}
    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), \sub{t}{\sigma} \\
    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
  \end{alignat*}
   the identities
  \begin{align*}
  \subTel{\Xi}{\sigma} & = \subTel{\Xi}{\tau}
                         & \subTy{A}{\sigma \upuparrows \Xi} & =^{\lam{\gamma}\Ty\;(\Gamma \dplus \subTel{\Xi}{\gamma})}_{\alert{\cc}} \subTy{A}{\tau \upuparrows \Xi} \\
  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus \subTel{\Xi}{\gamma})\;(\Delta\dplus\Xi)}_{\alert{\cc}} \tau \upuparrows \Xi
%    &
%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
  \end{align*}
  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
  We show the statement for $\pi_1\beta: \pi_1(\sigma, t) = t$ and the other three cases follow similarly.

@@

\propProvableFunc*
\begin{proof}
  TODO
\end{proof}

\equivalence*
\begin{proof}
  TODO
\end{proof}

>>>>

The lifting $\sigma^+$ interacts well with identities and composition.
For comparision, the reader is invited to prove the corresponding QIIT version with transports of the following lemma.
%
\begin{restatable}{lemma}{lemIdentities}\label{lem:lifting}
  For all contexts $\Gamma$, $\Delta$, $\Theta$, types $A : \Ty\,\Theta$ and substitutions $\sigma : \Sub\,\Gamma\,\Delta$ and $\tau : \Sub\,\Delta\,\Theta$, we have $\idS^+ =^{\Sub\;(\Theta, A)\;(\Theta, A)} \idS$ and $(\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+$.
\end{restatable}

@@

As we are working with a different type theory from \cref{subsec:SC+U}, we  have to again prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$, but the same argument still applies:
\begin{restatable}{proposition}{propCorrectnessThree}\label{prop:correctness-3}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
\end{restatable}

@@

\propConfluenceTwo*
\begin{proof}
  Termination is once again clear, since all recursion is structural.
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms
  \[
    \subTy{(\El\,u)}{\idS}, \qquad \subTy{(\El\,u)}{\sigma;\tau}, \qquad \subTy{(\El\,u)}{\pi_1(\sigma, t)}, \qquad\text{and}\qquad \subTy{(\El\,u)}{\pi_1(\sigma; \tau)}
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El\,(\subTm{\subTm{u}{\tau}}{\sigma}), \qquad \El\,(\subTm{u}{\sigma}), \qquad\text{and}\qquad
    \El\,(\subTm{\subTm{u}{\pi_1\tau}}{\sigma})
  \]
  respectively, no matter which function clause of $\subTy{\blank}{\blank}$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}

@@

\propCoherence*
\begin{proof}
  For term substitution, the claim follows by \cref{prop:correctness-1}: if
  $\cc : \sigma = \tau$ then
  \begin{align*}
    \subTm{t}{\sigma} & = \sub{t}{\sigma} && \text{by \cref{prop:correctness-1}} \\
                      & = \sub{t}{\tau} && \text{by congruence and $\cc$} \\
                      & = \subTm{t}{\tau} && \text{by \cref{prop:correctness-1}}
  \end{align*}
  For type substitution, by \cref{re:coherence-proof}, we only have to consider the following cases
  \begin{align*}
    \subTy{A}{\sigma; (\tau, t)}    & = \subTy{A}{(\sigma;\tau), \sub{t}{\sigma}},
                                     & \subTy{A}{\sigma}               & = \subTy{A}{\emptyctx},
                                     & \subTy{A}{\sigma}               & = \subTy{A}{\pi_1\sigma, \pi_2\sigma}
  \end{align*}
  while other cases hold strictly.
  Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
  However, all cases follow easily from \cref{prop:correctness-1}.
  For example, consider the first case:
  \begin{align*}
    \subTy{(\El\,u)}{\sigma; (\tau, t)} & \judgeq \El(\subTm{u}{\sigma; (\tau, t)}) && \text{by \eqref{eq:def-type-subst-7}} \\
%                                      & = \El( \sub{u}{\sigma; (\tau, t)})            && \text{by \cref{prop:correctness-1}} \\
%                                      & = \El( \sub{u}{\sigma; \tau, \sub{t}{\sigma}})     && \text{by $\mathsf{concat}$} \\
%                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by \cref{prop:correctness-1}} \\
                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by coherence of $\subTm{\blank}{\blank}$} \\
                                      & \judgeq \subTy{\left(\El\,u\right)}{\sigma; \tau, \sub{t}{\sigma}} && \text{by \eqref{eq:def-type-subst-7} }
  \end{align*}
  The other two cases are dealt with in exactly the same way.
\end{proof}

@@

\lemIdentities*
\begin{proof}
  We have
  \[
    \idS^+ \judgeq ((\pi_1\idS; \idS), \pi_2 \idS) = (\pi_1\idS, \pi_2 \idS) = \idS
  \]
  For commuting with composition, first note that $\pi_1(\sigma; \tau) = \sigma; \pi_1 \tau$ by $\mathsf{concat}$. In particular, we have $\sigma; \pi_1 \idS = \sigma^+; \pi_1\idS$. Now
  \begin{align*}
    (\sigma; \tau)^+ &= (\pi_1\idS; (\sigma; \tau), \pi_2 \idS) \\
                     &= ((\pi_1\idS; \sigma); \tau, \pi_2 \idS) \\
                     &= ((\sigma^+; \pi_1\idS); \tau, \sub{\pi_2 \idS}{\sigma^+}) \\
                     &= (\sigma^+; (\pi_1\idS; \tau), \pi_2 \idS) \\
                     &= \sigma^+; \tau^+
  \end{align*}
  as required.
\end{proof}

@@

\propCorrectnessThree*
\begin{proof}
  Again by induction on $\sigma$.
\end{proof}

\propCoherenceUparrow*
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3}.
\end{proof}

\coherenceTel*
\begin{proof}
  The proposition is easy to prove for the following equality
  constructors, since the lifting and substitution operations compute
  independently for these cases:
  \[
    \mathsf{idr} : \sigma;\idS = \sigma, \quad
    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma),
  \]
  Take $\mathsf{idr}$ as an example, we can see 
  \begin{align*}
    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows \subTel{\Xi}{\idS}) ; (\idS \upuparrows \Xi) \\
                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
                                   & = \sigma \upuparrows \Xi
  \end{align*}
  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.

@@

The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
  %\LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
  For the following equality constructors $\cc : \sigma = \tau$
  \begin{alignat*}{5}
    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), \sub{t}{\sigma} \\
    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
  \end{alignat*}
   the identities
  \begin{align*}
  \subTel{\Xi}{\sigma} & = \subTel{\Xi}{\tau}
                         & \subTy{A}{\sigma \upuparrows \Xi} & =^{\lam{\gamma}\Ty\;(\Gamma \dplus \subTel{\Xi}{\gamma})}_{\alert{\cc}} \subTy{A}{\tau \upuparrows \Xi} \\
  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus \subTel{\Xi}{\gamma})\;(\Delta\dplus\Xi)}_{\alert{\cc}} \tau \upuparrows \Xi
%    &
%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
  \end{align*}
  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
  We show the statement for $\pi_1\beta: \pi_1(\sigma, t) = t$ and the other three cases follow similarly.

@@

\propProvableFunc*
\begin{proof}
  These properties are proven mutually with the definition of the
  eliminators. The property $\elim_{\Ty}[]$ is proven by induction on the type $A$, while $\elim_{\Tm}[]$ and $\elim_{\Sub}[]$ are proven by induction on the substitution.
\end{proof}

\equivalence*
\begin{proof}
  See our \Agda formalisation for the exact definition of the functions back and forth, and how the roundtrips are identities.
\end{proof}
